---
title: "Data Cleaning and Analysis"
author: "Raquel Francisco"
date: "2024-03-12"
output: html_document
---
Load Packages
```{r setup, include=FALSE}
#Packages
library(here)
library(lattice)
library(ggplot2)
library(dplyr)
library(tidyverse)
library(tibble)
library(ggpubr)
library(epiR)
library(gridExtra)
```

## sVNT Expirement WV White-tailed deer
83 WTD in COVID Endemic Area
40 Juice Samples
40 Eluted Samples
34 Sera Samples


```{r all data}
getwd()
data<- read.csv(here('Data',"ELISA_2022Results.csv"))

str(data)

data$ID <- as.factor(data$ID)

tibble(data)
```

Seperate into 3 different data frames: Serum, Juice, Elution

Now will perform Optical Density comparisons!
Need to ensure that duplicates are correlated

```{r pressure, echo=FALSE}
sedata <- data %>% select(starts_with("s")) #serum
edata <- data %>% select(starts_with("e")) #elution
jdata <- data %>% select(starts_with("j")) #juice

#Serum
#not normal; right skewed
hist(sedata$sDup1)
hist(sedata$sDup2)

cor.test(sedata$sDup1, sedata$sDup2,
         method = "spearman")
#Highly correlated

#Elution
#not normal; right skewed
hist(edata$eDup1)
hist(edata$eDup2)

cor.test(edata$eDup1, edata$eDup2,
         method = "spearman")
#Correlated

#Juice
#more normal?
hist(jdata$jDup1)
hist(jdata$jDup2)

cor.test(jdata$jDup1, jdata$jDup2,
         method = "spearman")
#Highly correlated
```
Summary: Consistency of the assay with serum appears to be highest, then juice, last is elution.(Will use correlation co-efficent as justification for reasoning)

Will now compare Inhibition between duplicates, which is an output of OD, thus these are not independent.
```{r}
##SERUM
#more normal then ODs; left skewed
hist(sedata$sD1Inh)
hist(sedata$sD2Inh)

cor.test(sedata$sD1Inh, sedata$sD2Inh,
         method = "spearman")
#Highly correlated

ggscatter(sedata, x = "sD1Inh", y = "sD1Inh", 
          add = "reg.line", conf.int = TRUE, 
          cor.coef = TRUE, cor.method = "spearman",
          xlab = "Duplicate 1 Inhibition", ylab = "Duplicate 1 Inhibition")

##JUICE
#normalish
hist(jdata$jD1Inh)
hist(jdata$jD2Inh)

cor.test(jdata$jD1Inh, jdata$jD2Inh,
         method = "spearman")
#Highly correlated

ggscatter(jdata, x = "jD1Inh", y = "jD1Inh", 
          add = "reg.line", conf.int = TRUE, 
          cor.coef = TRUE, cor.method = "spearman",
          xlab = "Duplicate 1 Inhibition", ylab = "Duplicate 1 Inhibition")

##ELUTED WHOLE BLOOD
#not normal; left skewed
hist(edata$eD1Inh)
hist(edata$eD2Inh)

cor.test(edata$eD1Inh, edata$eD2Inh,
         method = "spearman")

ggscatter(edata, x = "eD1Inh", y = "eD1Inh", 
          add = "reg.line", conf.int = TRUE, 
          cor.coef = TRUE, cor.method = "spearman",
          xlab = "Duplicate 1 Inhibition", ylab = "Duplicate 1 Inhibition")

```
Summary: results similar to testing with corr between the OD, however may be easier to digest if data reported as Inhibition because that's how results will be reported.

##Correlations between biological sample types
#Serum vs Juice
```{r}
cor.test(data$sAvgInh, data$jAvgInh,
         method = "spearman")
#not as correlated

plotSJ <- ggscatter(data, x = "sAvgInh", y = "jAvgInh", 
          add = "reg.line", conf.int = TRUE, 
          cor.coef = TRUE, cor.method = "spearman",
          xlab = "Average Serum Inhibition", ylab = "Average Juice Inhibition")
plotSJ

```
#Serum vs Elution
```{r}
cor.test(data$sAvgInh, data$eAvgInh,
         method = "spearman")
#S = 5578, p-value = 0.4028
#alternative hypothesi: true rho is not equal to 0
#sample estimates:rho 0.1477464 

plotSE <- ggscatter(data, x = "sAvgInh", y = "eAvgInh", 
          add = "reg.line", conf.int = TRUE, 
          cor.coef = TRUE, cor.method = "spearman",
          xlab = "Average Serum Inhibition", ylab = "Average Elution Inhibition")

plotSE
#note gray is 95% confidence interval
#grid.arrange(plotSE, plotSJ, nrow = 1)
```
##Manually creating tables: 
#Sensitivity, Specificity, PPV, NPV
```{r}
count(data, vars= 'sCodeAvg')
sum(data$sCodeAvg)
#Tables made manually and double checked in epiR

##building table for Elution
elTable <- matrix(c(2,2,13,17), ncol=2, byrow=TRUE)
colnames(elTable) <- c("T-", "T+")
rownames(elTable) <- c("Ab-", "Ab+")
elTable            

margin1 <- margin.table(elTable, margin = 1)
margin1
margin2 <- margin.table(elTable, margin = 2)
margin2

Abpos <- margin1[2]
Abpos
Abneg <- margin1[1]
Abneg
testpos <- margin2[2]
testpos
testneg <- margin2[1]
testneg
truepos <- elTable[2, 2]
truepos
falsepos <- elTable[1, 2]
falsepos
trueneg <- elTable[1, 1]
trueneg
falseneg <- elTable[2, 1]
falseneg

#sensitivity
truepos/Abpos
#0.5666667 

#specificity
trueneg/Abneg
#0.5

#PPV
truepos/testpos
#0.8947368 

#NPV
trueneg/testneg
#0.1333333 

##building table for Juice
jTable <- matrix(c(0,4,4,26), ncol=2, byrow=TRUE)
colnames(jTable) <- c("T-", "T+")
rownames(jTable) <- c("Ab-", "Ab+")
jTable            

jmargin1 <- margin.table(jTable, margin = 1)
jmargin1
jmargin2 <- margin.table(jTable, margin = 2)
jmargin2

jAbpos <- jmargin1[2]
jAbpos
jAbneg <- jmargin1[1]
jAbneg
jtestpos <- jmargin2[2]
jtestpos
jtestneg <- jmargin2[1]
jtestneg
jtruepos <- jTable[2, 2]
jtruepos
jfalsepos <- jTable[1, 2]
jfalsepos
jtrueneg <- jTable[1, 1]
jtrueneg
jfalseneg <- jTable[2, 1]
jfalseneg

#sensitivity
jtruepos/jAbpos
#0.8666667 

#specificity
jtrueneg/jAbneg
#0

#PPV
jtruepos/jtestpos
#0.8666667 

#NPV
jtrueneg/jtestneg
#0
```


##epiR package
#Confirmation
```{r}
#Elution vs Serum
elTable
#edit table to flip order
elTable2 <- matrix(c(17,2,13,2), ncol=2, byrow=TRUE)
colnames(elTable2) <- c("T+", "T-")
rownames(elTable2) <- c("Ab+", "Ab-")
elTable2

rval<- epi.tests(elTable2, conf.level = 0.95)
print(rval)
str(rval)
rval$detail

#Juice vs Serum
jTable2 <- matrix(c(26,4,4,0), ncol=2, byrow=TRUE)
colnames(jTable2) <- c("T+", "T-")
rownames(jTable2) <- c("Ab+", "Ab-")
jTable2

rval2<- epi.tests(jTable2, conf.level = 0.95)
print(rval2)
rval2$detail
```

Create a df from plot
```{r}
edata$Type <- ("Elution")
sedata$Type <- ("Serum")
jdata$Type <- ("Juice")

plote <- edata %>% select('eAvgInh'|'eAvgResult'|'Type')
plote <- plote %>% rename('AvgInh' = 'eAvgInh') %>% rename('AvgResult' = 'eAvgResult')

plots <- sedata %>% select('sAvgInh'|'sAvgResult'|'Type')
plots <- plots %>% rename('AvgInh' = 'sAvgInh') %>% rename('AvgResult' = 'sAvgResult')

plotj <- jdata %>% select('jAvgInh'|'jAvgResult'|'Type')
plotj <- plotj %>% rename('AvgInh' = 'jAvgInh') %>% rename('AvgResult' = 'jAvgResult')

plotdata <- bind_rows(plote, plots, plotj)
tibble(plotdata)

```

#Create Boxplot with raw points

```{r}
pplot <- ggplot(plotdata, aes(x = Type, y = AvgInh, color = AvgResult)) + 
  geom_boxplot() +
  geom_point(position=position_dodge(width=0.75)) + 
  geom_hline(yintercept=30, linetype="dashed", size=1) 
 

pplot

pplot + 
  ggtitle('Presence of Neutralizing Antibodies in White-tailed Deer Biological Samples') + 
  scale_x_discrete(labels = c('Elution','Serosanguineous Fluid','Serum')) +
  ylab('Average Inhibition') + 
  xlab('') + 
  labs(color='') +
  theme(text = element_text(family = "serif"),
        plot.title = element_text(size = 16), 
        axis.text = element_text(size = 14),
        legend.text = element_text(size = 14),
        axis.title = element_text(size = 14),
        panel.background = element_rect(fill = "white", colour = "grey50"))
```

